Normalization Circularity in Population-Based FA Analysis

### 2011-06-06 ANTsTBSS/paper.tex
ENH: compile script + revised abstract
ENH: try to explain some of the issues --- edited to before section 2.

\begin{abstract}
Template-based normalization plays a critical role in many population
studies based on diffusion tensor imaging (DTI).  Several current
approaches compute the fractional anisotropy (FA) from the DTI and then
directly register the FA image to a FA template before performing
statistical comparison between different groups within a population.  In
this paper, we demonstrate that this analysis strategy induces selection
bias that varies in strength with the similarity metric used in
registration.  Specifically, we show that the popular sum of squared
difference intensity metric maximizes effect size, rather than
anatomical alignment, and consistently overestimates statistical
significance.  Consequently, we advocate a more traditional approach
that normalizes a population of FA images through the associated T1
anatomical images.  We show that the Symmetric Group Normalization
strategy for unbiased template construction from anatomical (e.g.
T1-weighted) images provides a valid and unbiased coordinate system for
statistical inference of FA differences and yet retains detection power
in comparison to FA-based normalization methods.  We establish this
outcome by using open data in normal controls from the International
Neuroimaging Data-sharing Initiative as well as a cohort of 16 traumatic
brain injury survivors compared with 17 control subjects.  Alignment
comparisons with the normalization component of the popular tract-based
spatial statistics (TBSS) show that our anatomically-based template
approach produces good alignments despite the absence of any direct
FA-to-FA registration.  Of practical consideration we note that the
software tools discussed in this work are provided to the public as part
of the open source Advanced Normalization Tools (ANTs) repository.

@@

The seminal work of Basser et al. \cite{Basser1994a,Basser1994} established diffusion tensor imaging (DTI) as a viable investigatory MRI technique.  Continued quantitative development has furthered its applicability to probing clinical questions of interest particularly pertaining to the microstructure of white matter \cite{Basser1996,Assaf2008}.  Given this sensitivity to brain architecture, DTI has found multiple uses in assessing neuro-structural differences in cross-population studies \cite{Kubicki2005,Arnone2008,Kantarci2010,Rametti2010}
including subjects with traumatic brain injury (TBI) \cite{Caeyenberghs2010,Jiang2010,Warner2010}.

Normalization (also known as image registration) maps each image in the population to a common coordinate system thus enabling statistical comparison across patient groups or association with other variables such as genetics, cognition or age.

Two popular comprehensive software packages for assessing population FA differences include 
the SPM Matlab-based toolkit%
\footnote{
http://www.fil.ion.ucl.ac.uk/spm/
}
for voxel-based morphometry (VBM)
and the tract-based spatial statistics (TBSS) framework \cite{Smith2006}.%
\footnote{
http://www.fmrib.ox.ac.uk/fsl/tbss/index.html
} 
Although concerns regarding the general 
application of VBM \cite{Bookstein2001,Ashburner2001,Davatzikos2004}
and its application to FA data \cite{Jones2005,Chung2008} has been discussed 
in the literature, it continues to find application in FA population 
studies \cite{Kakeda2010,Takao2010}.
The TBSS framework was partially developed in response to these concerns
in which voxel values are projected onto a template white matter skeleton for 
increased statistical power.  
%(e.g. \cite{Arnone2008,Bodini2009}).

Although there are substantive differences in normalization schemes
between these and other image registration approaches, a core
commonality includes direct FA-to-FA registration.
Potential difficulties associated with direct FA-to-FA normalization is exemplified by 
a standard population study in which all subject FA images are mapped to a representative 
template using the popular sum of squared differences (SSD) image similarity metric.%
\footnote{
In addition to TBSS and SPM, other popular packages in the neuroimaging community 
which use the SSD metric includes AIR \cite{Woods1998} and LDDMM \cite{Beg2005}.
}  
Given a set of $M$ subject images, $\{I_1, I_2, \ldots, I_M\}$, and
a template image, $J$, population normalization concerns finding the set of $M$ transformations, 
$\boldsymbol{\mathcal{T}} = \{\mathcal{T}_1, \mathcal{T}_2,\ldots,\mathcal{T}_M \}$, which minimize the SSD metric over the population, i.e.
\begin{align}
  \boldsymbol{\mathcal{T}} = \underset{\mathcal{T}_1, \ldots, \mathcal{T}_M}{\operatorname{argmin}}
    \sum_{m=1}^M\sum_{n=1}^N \left( I^n_m( \mathcal{T}_m(\mathbf{x}) ) - J \right)^2.
\end{align}
Switching the order of the summations (since the transformations are mutually independent) and 
recognizing that a principal criterion for the selection of $J$ is such that it be a good approximation of the 
mean of the aligned images, it becomes apparent that the transformation solution
\begin{align}\label{eq:variance}
  \boldsymbol{\mathcal{T}} = \underset{\mathcal{T}_1, \ldots, \mathcal{T}_M}{\operatorname{argmin}}
    \sum_{n=1}^N \underbrace{\sum_{m=1}^M \left( I^n_m( \mathcal{T}_m(\mathbf{x}) ) - J \right)^2}_{
    \propto\mathrm{\,voxelwise\,variance}}
\end{align}
is that which minimizes the voxelwise variance or, equivalently,
maximizes various voxelwise statistical quantities including the common Student's $t$-test.%
Note that many of the standard similarity metrics are also going to exhibit this
effect to some degree as they minimize some local measure of intensity-based distance 
between the subject and the template.

Thus, instead of normalization based on anatomical alignment followed by
statistical testing, direct FA normalization 
explicitly conflates anatomical alignment with optimizing certain statistical testing results
evoking a sense of circularity in the analysis \cite{Kriegeskorte2010}.
According to Kriegeskrote et al., ``An analysis is circular (or nonindependent) if it is based on
data that were selected for showing the effect of interest or a related effect.''  As illustrated with
Eqn. \ref{eq:variance}, direct FA-to-FA registration causes voxels to be matched (or selected) according to the
the statistical testing result (or the effect of interest) instead of anatomical alignment.
%This results in a significant increase in the number of false positives for this type
%of testing which we illustrate with the following example.

@@

Axial slice number 159 was extracted from \verb#subject_04# of the discrete BrainWeb database
\cite{Aubert-Broche2006} which is shown in Fig. \ref{fig:simulated_data}.  After thresholding out 
the white matter and assigning all white matter voxels a value of 0.7, 10\% of these white 
matter voxels were randomly selected to be assigned a value of 0.6 in the simulated control 
template and a value of 0.3 in the simulated patient template.  
Note that these values were arbitrarily selected to be more or less in the range of typical
FA values.  Gaussian noise (mean = 0.0, standard deviation = 0.1) was added to the white
matter template followed by a Gaussian smoothing (standard deviation = 1.5 mm) shown in Fig. \ref{fig:simulated_data}(b).  A simulated control cohort was created consisting of 20 subjects.  
Starting with the simulated control template, each of the 20 subjects is created by adding
Gaussian noise (mean = 0.0, standard deviation = 0.1) followed by 
Gaussian smoothing (standard deviation = 1.5 mm).  A simulated patient cohort of 20 subjects
was created similarly using the simulated patient template.  A sample control subject and sample
patient subject are given, respectively, in Fig. \ref{fig:simulated_data}(c) and (d).

\begin{figure*}
\begin{center}
\begin{tabular}{c}
  \includegraphics[width=180mm]{simulatedTesting.pdf}
\end{tabular}
\caption{Student's $t$-test and the nonparametric Wilcoxon rank-sum test results (one-tailed, $\alpha = 0.95$) for the simulated experiment adjusted for multiple comparisons (FDR).
The true significance curves are drawn in black.  Elevated statistical power is seen with all 4 
         metrics for both tests although the effect is less in the more robust Wilcoxon results  Not shown are the significance curves for the Brunner-Munzel testing \cite{Rorden2007} which showed similar effects to the results on the left.
        }
\label{fig:simulated_testing}
\end{center}        
\end{figure*}

After creation of the simulated data set, we performed voxelwise testing (both $t$-testing
and Wilcoxon rank-sum testing, one-tailed, $\alpha = 0.95$) with false discovery
rate correction within the white matter to get ``true'' statistical significance 
curves.  These curves are plotted in black in Fig. \ref{fig:simulated_testing}.  To illustrate the 
effect of similarity metric on statistical significance, we performed ANTs registration
\cite{Avants2011} of each subject to the template
(Fig. \ref{fig:simulated_data}(b)) using cross correlation (CC), mutual information (MI), 
Demons \cite{Thirion1998}, and sum of squared differences (SSD). A representative deformed image
using the SSD metric and its corresponding deformed grid is given in 
Fig. \ref{fig:simulated_data_warped}. 
The resulting statistical significance curves are plotted with the true significance curves in Fig. 
\ref{fig:simulated_testing}.

This example demonstrates the undesirable phenomenon previously described.  Even though these
images were anatomically aligned prior to registration, the resulting deformations and 
increased false statistical power can be explained in the context of Eqn. (\ref{eq:variance}).  
Common similarity metric functions can induce false positives by minimizing local intensity 
differences between the subjects and template which explicitly maximizes statistical
testing results in contrast to maximizing anatomical alignment.  This effect is strongest in 
the SSD and Demons metrics and weakest in the MI and CC metrics likely due
to the degree of correspondence to Eqn. (\ref{eq:variance}).   This is also supported by contrasting
the $t$-test results given in Fig. \ref{fig:simulated_testing}(a) with the non-parametric Wilcoxon
rank-sum testing (Fig. \ref{fig:simulated_testing}(b)) where as large of an effect for the latter would not be expected although an effect is
still apparent.

Although various processing choices mitigate these effects, such as Gaussian smoothing, nonparametric
testing, and
the use of more robust similarity metrics, such choices are often made on an ad hoc basis.  
In contrast, in this work we strongly advocate the general application
%A more principled approach is the use of an anatomical, i.e. T1-weighted, unbiased shape/intensity 
%template constructed from the population images.  DTI-derived scalar images can then be mapped 
%to their corresponding T1-weighted image in a highly constrained manner and subsequently 
%mapped to the template via nonrigid registration of the T1-weighted image.  This minimizes 
%possible complications associated with direct FA-to-FA image registration.
%We recognize that using the anatomical images in the alignment process has been previously proposed (in fact, such a suggestion occurs in the {\it Future directions} section of the original TBSS paper).  
of a high performance \cite{Klein2009}, nonrigid registration algorithm \cite{Avants2011}, which forms the underlying machinery for an optimal template construction algorithm \cite{Avants2010}, in facilitating alignment between DTI-derived scalar images for cross-population studies.

>>>>

\begin{abstract}
Template-based normalization plays a critical role in many population
studies based on diffusion tensor imaging (DTI).  Several current
approaches compute the fractional anisotropy (FA) from the DTI and then
directly register the FA image to a FA template before performing
statistical comparison between different groups within a population.  In
this paper, we demonstrate that this analysis strategy induces selection
bias that varies in strength with the similarity metric used in
registration.  Specifically, we show that the popular sum of squared
difference intensity metric maximizes effect size, rather than
anatomical alignment, and consistently overestimates statistical
significance.  Consequently, we advocate a more traditional approach
that normalizes a population of FA images through the associated T1
anatomical images.  We show that the Symmetric Group Normalization
strategy for unbiased template construction from anatomical (e.g.
T1-weighted) images provides a valid and unbiased coordinate system for
statistical inference of FA differences and yet retains detection power
in comparison to FA-based normalization methods.  We establish this
outcome by using open data in normal controls from the International
Neuroimaging Data-sharing Initiative as well as a cohort of 16 traumatic
brain injury survivors compared with 17 control subjects.  Alignment
comparisons with the normalization component of the popular tract-based
spatial statistics (TBSS) show that our anatomically-based template
approach produces good alignments despite the absence of any direct
FA-to-FA registration.  Of practical consideration we note that the
software tools discussed in this work are provided to the public as part
of the open source Advanced Normalization Tools (ANTs) repository.  In conclusion, we recommend against circular analysis strategies that select image similarity metrics that are tied directly to the ultimate statistical functions used to assess population effects.

@@

The seminal work of Basser et al. \cite{Basser1994a,Basser1994} established diffusion tensor imaging (DTI) as a viable investigatory MRI technique.  DTI's sensitivity to brain architecture \cite{Basser1996,Assaf2008} enables multiple uses in assessing neuro-structural differences in cross-population studies \cite{Kubicki2005,Arnone2008,Kantarci2010,Rametti2010} including subjects with traumatic brain injury (TBI) \cite{Caeyenberghs2010,Jiang2010,Warner2010}.  Two popular comprehensive software packages for assessing population FA differences include the SPM Matlab-based toolkit%
\footnote{
http://www.fil.ion.ucl.ac.uk/spm/
}
for voxel-based morphometry (VBM)
and the tract-based spatial statistics (TBSS) framework \cite{Smith2006}.%
\footnote{
http://www.fmrib.ox.ac.uk/fsl/tbss/index.html
} 
VBM\cite{Ashburner2001} continues to find application to FA studies\cite{Kakeda2010,Takao2010} despite concerns about its validity\cite{Bookstein2001,Davatzikos2004} in DTI \cite{Jones2005,Chung2008}.  The TBSS framework was partially developed in response to these concerns in which voxel values are projected onto a template white matter skeleton for increased statistical power.  
%(e.g. \cite{Arnone2008,Bodini2009}).

Although there are substantive differences in normalization schemes
between these and other image registration approaches, a core
commonality includes direct FA-to-FA registration.  The sum of squared differences (SSD) image similarity metric is perhaps the easiest to interpret (it drives the image difference to zero), is computationally efficient (just compute the image difference and an image gradient) and is therefore widely used.  Some of the most popular registration methods (SPM, Demons, FSL's FNIRT) rely on this metric and provide reasonable performance levels \cite{Klein2009}.
\footnote{
In addition to TBSS and SPM, other popular packages in the neuroimaging community 
which use the SSD metric includes AIR \cite{Woods1998} and LDDMM \cite{Beg2005}.
}  
Now, suppose the SSD metric is optimized to find a set of $M$ transformations that map a population of $M$ FA images, $\{I_1, I_2, \ldots, I_M\}$, to a representative FA template, $J$. Then, the algorithm will compute transformations
$\boldsymbol{\mathcal{T}} = \{\mathcal{T}_1, \mathcal{T}_2,\ldots,\mathcal{T}_M \}$, which minimize the SSD metric over the population, i.e.
\begin{align}
  \boldsymbol{\mathcal{T}} = \underset{\mathcal{T}_1, \ldots, \mathcal{T}_M}{\operatorname{argmin}}
    \sum_{m=1}^M\sum_{n=1}^N \left( I^n_m( \mathcal{T}_m(\mathbf{x}) ) - J \right)^2.
\end{align}
Switching the order of the summations (since the transformations are mutually independent) and recognizing that a principal criterion for the selection of $J$ is such that it be a good approximation of the mean of the aligned images, it becomes apparent that the transformation solution
\begin{align}\label{eq:variance}
  \boldsymbol{\mathcal{T}} = \underset{\mathcal{T}_1, \ldots, \mathcal{T}_M}{\operatorname{argmin}}
    \sum_{n=1}^N \underbrace{\sum_{m=1}^M \left( I^n_m( \mathcal{T}_m(\mathbf{x}) ) - J \right)^2}_{
    \propto\mathrm{\,voxelwise\,variance}}
\end{align}
is that which minimizes the voxelwise variance.  Reducing variance will also directly affect the value of a standard population statistic, the Student's $t$-test.  Other standard similarity metrics will also exhibit this
trend to some degree as they minimize some local measure of intensity-based distance between the subject and the template.

Thus, instead of normalization based on anatomical alignment followed by
statistical testing, direct FA normalization explicitly conflates
anatomical alignment with optimizing the very statistical testing
results that one will ultimately use to assess hypotheses.  This immediately evokes a sense of circularity in the analysis \cite{Kriegeskorte2010}.
According to Kriegeskrote et al., ``An analysis is circular (or nonindependent) if it is based on data that were selected for showing the effect of interest or a related effect.''  As illustrated with
Eqn. \ref{eq:variance}, direct FA-to-FA registration will find transformations that align (or select) voxels such that the statistical testing result (or the effect of interest) is increased (assuming one is using Student's test).  While the SSD metric is able to provide visually attractive alignments across subjects, the voxel-level details (that is, which voxels are matched where) are, as we show in the following simulation study, strikingly important in terms of the effect on detection power.

%This results in a significant increase in the number of false positives for this type
%of testing which we illustrate with the following example.

@@

Axial slice number 159 was extracted from \verb#subject_04# of the discrete BrainWeb database
\cite{Aubert-Broche2006} which is shown in Fig. \ref{fig:simulated_data}.  After thresholding out 
the white matter and assigning all white matter voxels a value of 0.7, 10\% of these white 
matter voxels were randomly selected to be assigned a value of 0.6 in the simulated control 
template and a value of 0.3 in the simulated patient template.  
Note that these values were arbitrarily selected to be more or less in the range of typical
FA values.  Gaussian noise (mean = 0.0, standard deviation = 0.1) was added to the white
matter template followed by a Gaussian smoothing (standard deviation = 1.5 mm) shown in Fig. \ref{fig:simulated_data}(b).  A simulated control cohort was created consisting of 20 subjects.  
Starting with the simulated control template, each of the 20 subjects is created by adding
Gaussian noise (mean = 0.0, standard deviation = 0.1) followed by 
Gaussian smoothing (standard deviation = 1.5 mm).  Thus, each subject is differentiated only by a unique random draw from the noise distribution and there is no real shape difference between subjects.  A simulated patient cohort of 20 subjects
was created similarly using the simulated patient template.  A sample control subject and sample
patient subject are given, respectively, in Fig. \ref{fig:simulated_data}(c) and (d).

\begin{figure*}
\begin{center}
\begin{tabular}{c}
  \includegraphics[width=180mm]{simulatedTesting.pdf}
\end{tabular}
\caption{Student's $t$-test and the nonparametric Wilcoxon rank-sum test results (one-tailed, $\alpha = 0.95$) for the simulated experiment adjusted for multiple comparisons (FDR).
The true significance curves are drawn in black.  Elevated statistical power is seen with all 4 
         metrics for both tests although the effect is less in the more robust Wilcoxon results. Not shown are the significance curves for the Brunner-Munzel testing \cite{Rorden2007} which showed similar effects to the results on the left.
        }
\label{fig:simulated_testing}
\end{center}        
\end{figure*}

After creation of the simulated data set, we performed voxelwise testing (both $t$-testing
and Wilcoxon rank-sum testing, one-tailed, $\alpha = 0.95$) with false discovery
rate correction within the white matter to get ``true'' cumulative distribution functions (cdfs).  The cdf vertical axis shows the cumulative number of voxels with corrected p-values at or below a given significance level (horizontal axis) and thus may be used as a surrogate measure for spatial sensitivity.  From here, we denote these graphs as "statistical significance curves."  These curves are plotted in black in Fig. \ref{fig:simulated_testing}.  To illustrate the 
effect of similarity metric on statistical significance, we performed ANTs registration \cite{Avants2011} of each subject to the template
(Fig. \ref{fig:simulated_data}(b)) using cross correlation (CC), mutual information (MI), 
Demons \cite{Thirion1998}, and sum of squared differences (SSD). A representative deformed image
using the SSD metric and its corresponding deformed grid is given in 
Fig. \ref{fig:simulated_data_warped}. 
The resulting statistical significance curves are plotted with the true significance curves in Fig. 
\ref{fig:simulated_testing}.

This example demonstrates the undesirable phenomenon previously described.  Even though these images were anatomically aligned prior to registration, the resulting deformations (falsely) increased detection power.  This effect can be explained in the context of Eqn. (\ref{eq:variance}) and by the fact that we added noise to the underlying distribution which allows the deformable mapping algorithms the freedom to transform the positions of the most "powerful" voxels to minimize variance across the population (and increase the t-test).  Note, however, that this effect is increased by the fact that noise was added to the data. {\textcolor{red}{FIXME --- does more / less noise increase / decrease the exaggerated power? }}
Common similarity metric functions can therefore induce false positives by minimizing local intensity 
variance between the subjects and template which explicitly maximizes statistical
testing results in contrast to maximizing anatomical alignment.  This effect is strongest in 
the SSD and Demons metrics and weakest in the MI and CC metrics likely due
to the degree of correspondence to Eqn. (\ref{eq:variance}).   This is also supported by contrasting
the $t$-test results given in Fig. \ref{fig:simulated_testing}(a) with the non-parametric Wilcoxon
rank-sum testing (Fig. \ref{fig:simulated_testing}(b)).  The rank test does not explicitly use variance in its assessment and thus exhibits far less statistical exaggeration.

Although various processing choices mitigate these effects, such as Gaussian smoothing, nonparametric testing, and the use of more robust similarity metrics, such choices are often made on an ad hoc basis.  
In contrast, in this work we strongly advocate the general application
%A more principled approach is the use of an anatomical, i.e. T1-weighted, unbiased shape/intensity 
%template constructed from the population images.  DTI-derived scalar images can then be mapped 
%to their corresponding T1-weighted image in a highly constrained manner and subsequently 
%mapped to the template via nonrigid registration of the T1-weighted image.  This minimizes 
%possible complications associated with direct FA-to-FA image registration.
%We recognize that using the anatomical images in the alignment process has been previously proposed (in fact, such a suggestion occurs in the {\it Future directions} section of the original TBSS paper).  
of a high performance \cite{Klein2009}, nonrigid registration algorithm \cite{Avants2011}, which forms the underlying machinery for an optimal template construction algorithm \cite{Avants2010}, in facilitating alignment between DTI-derived scalar images for cross-population studies.  The key point, here, is that the FA images are mapped to the common reference space via intra-subject DT-to-T1 mapping followed by T1-to-T1 alignment.  The T1-to-T1 step is (statistically) independent of the FA values which are the ultimate statistical target of hypothesis testing in FA population studies.  Below, we show that this same pattern of effects is reproduced in real data and detail our proposed solution.
